Factors affecting health sector expenditure in OECD countries

### 2025-04-04 code/analysis-code/statistical-analysis.qmd
changes to regression tables

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(broom)      
library(kableExtra) 
library(webshot)
library(magick)
library(here)
library(modelsummary)
library(plm)
library(lmtest)
library(gt)
library(webshot2)
library(fixest)
library(lubridate)
library(Metrics)

@@

@@

modelsummary(
  model,
  output = here("results", "tables", "preliminary_regression_results.png"),
  title = "This table represents the preliminary regresssion model",
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  notes = list(
    "Standard errors are shown in parentheses.",
    "Significance levels: * p < 0.1, ** p < 0.05, *** p < 0.01."
  )
)

@@

 ADF test (combined)
```{r}
library(plm)

# 1. Prepare panel data (handle NAs first)
train_clean <- na.omit(train[, c("CountryName", "Year", "Current_health_expenditure_per_capita_current_US")])

# 2. Check time dimensions
time_counts <- table(train_clean$country)
print(time_counts)

# 3. Run Fisher-type test with reduced lags
pdata <- pdata.frame(train_clean, index = c("CountryName", "Year"))
fisher_test <- purtest(pdata$Current_health_expenditure_per_capita_current_US,
                      test = "madwu",
                      exo = "intercept",
                      lags = 1) # Force shorter lags

```{r}
library(kableExtra)
library(here)

# 1. Create properly formatted results table
unit_root_table <- data.frame(
  Test = "Fisher-Type Panel Unit Root Test (ADF)",
  Statistic = fisher_test$statistic$statistic,
  DF = fisher_test$statistic$parameter,
  `p-value` = fisher_test$statistic$p.value,
  check.names = FALSE
)

# 2. Create and save HTML table
unit_root_table %>%
  kbl(format = "html", digits = 4, align = "c",
      caption = "Panel Unit Root Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  add_footnote("Null hypothesis: All series contain unit roots", 
              notation = "none") %>%
  save_kable(here("results", "tables", "unit_root_results.html"))
```

@@

# Option 1: Simple version with try()
adf_results <- by(train, train$CountryName, function(x) {
  try(adf.test(na.omit(x$Current_health_expenditure_per_capita_current_US)),
  silent = TRUE)
})

# Option 2: More robust version with error handling
adf_results <- lapply(split(train, train$CountryName), function(df) {
  if (sum(!is.na(df$Current_health_expenditure_per_capita_current_US)) >= 3) {
    tryCatch({
      adf.test(na.omit(df$Current_health_expenditure_per_capita_current_US))
    }, error = function(e) {
      return(list(statistic = NA, p.value = NA, 
                 note = paste("Error:", e$message)))
    })
  } else {
    return(list(statistic = NA, p.value = NA, 
               note = "Not enough observations"))
  }
})

@@

>>>>

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(broom)      
library(kableExtra) 
library(webshot)
library(magick)
library(here)
library(modelsummary)
library(plm)
library(lmtest)
library(gt)
library(webshot2)
library(fixest)
library(lubridate)
library(Metrics)
library(kableExtra)
library(here)

@@

model1 <- lm(Current_health_expenditure_per_capita_current_US ~ 
              GDP_growth_annual_ + 
              Income_share_held_by_lowest_20 + 
              Control_of_Corruption_Estimate + 
              Life_expectancy_at_birth_total_years + 
Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate + 
              Trade__of_GDP+ Foreign_direct_investment_net_inflows_BoP_current_US+
  Access_to_electricity_rural__of_rural_population +
  Age_dependency_ratio__of_workingage_population +
  Military_expenditure__of_GDP +
  Literacy_rate_adult_total__of_people_ages_15_and_above

  , 
            data = train)

modelsummary(
  model1,
  output = here("results", "tables", "preliminary_regression_results.png"),
  title = "This table represents the preliminary regresssion model",
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  notes = list(
    "Standard errors are shown in parentheses.",
    "Significance levels: * p < 0.1, ** p < 0.05, *** p < 0.01."
  )
)

However, due to the missing values in the independent variables the total number of observation used in the regression is too low. So, we will use such variables which are available as much as possible for all countries.

```{r}
# primary regression model

@@

modelsummary(
  model,
  output = here("results", "tables", "primary_regression_results.png"),
  title = "This table represents the primary regresssion model",
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  notes = list(
    "Standard errors are shown in parentheses.",
    "Significance levels: * p < 0.1, ** p < 0.05, *** p < 0.01."
  )
)

@@

ADF test (combined)

```{r}
library(plm)

#handling NAs first
train_clean <- na.omit(train[, c("CountryName", "Year", "Current_health_expenditure_per_capita_current_US")])

# time dimensions
time_counts <- table(train_clean$country)
print(time_counts)

# Fisher-type test with reduced lags
pdata <- pdata.frame(train_clean, index = c("CountryName", "Year"))
fisher_test <- purtest(pdata$Current_health_expenditure_per_capita_current_US,
                      test = "madwu",
                      exo = "intercept",
                      lags = 1) # Force shorter lags

```
ADF test (combined) continued....

# 1. Creating properly formatted results table
unit_root_table <- data.frame(
  Test = "Fisher-Type Panel Unit Root Test (ADF)",
  Statistic = fisher_test$statistic$statistic,
  DF = fisher_test$statistic$parameter,
  `p-value` = fisher_test$statistic$p.value,
  check.names = FALSE
)

# creating and save HTML table
unit_root_table %>%
  kbl(format = "html", digits = 4, align = "c",
      caption = "Primary model Unit Root Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  add_footnote("Null hypothesis: All series contain unit roots", 
              notation = "none") %>%
  save_kable(here("results", "tables", "unit_root_results.html"))
```

@@

# adf result
adf_results <- lapply(split(train, train$CountryName), function(df) {
  if (sum(!is.na(df$Current_health_expenditure_per_capita_current_US)) >= 3) {
    tryCatch({
      adf.test(na.omit(df$Current_health_expenditure_per_capita_current_US))
    }, error = function(e) {
      return(list(statistic = NA, p.value = NA, 
                 note = paste("Error:", e$message)))
    })
  } else {
    return(list(statistic = NA, p.value = NA, 
               note = "Not enough observations"))
  }
})

@@

We will also test another version of our primary model excluding life expectancy to test if the results differ.

```{r}
# primary regression model

model2 <- lm(Current_health_expenditure_per_capita_current_US ~ 
              GDP_growth_annual_ + 
              Income_share_held_by_lowest_20 + 
              Control_of_Corruption_Estimate + 
Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate + 
              Trade__of_GDP+ Foreign_direct_investment_net_inflows_BoP_current_US+
  Access_to_electricity_rural__of_rural_population +
  Age_dependency_ratio__of_workingage_population
  , 
            data = train)

modelsummary(
  model2,
  output = here("results", "tables", "primary_regression_excl_LE.png"),
  title = "This table represents the primary regresssion model without life expectancy",
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  notes = list(
    "Standard errors are shown in parentheses.",
    "Significance levels: * p < 0.1, ** p < 0.05, *** p < 0.01."
  )
)

We can observe that the estimates and standard error don't change much even after the exclusion of life expectancy at birth, so our suspicion of endogeniety might not be that serious.

### 2025-04-04 code/eda-code/eda.qmd
changes to regression tables

```
We can not see a visible relation between these two factors.  Let's try it for France now

@@

# Combining both plots side by side
plot_pop_growth + plot_health_expenditure
```
We can not observe a distinct relation between the two variable, we can observe that the Health Expenditure (% of GDP) variable also show less fluctuation across the years for multiple countries. Lets see if Health Expenditure Per Capita (USD) over the Years shows any reation with annual population growth.

@@

```
The Health Expenditure Per Capita (USD) varriable in fact shows more variation and also show some connections with annual population growth.
```{r}

>>>>

We can not see a visible relation between these two factors.  Let's try it for France now

@@

# Combining both plots side by side
plot_pop_growth + plot_health_expenditure
```

We can not observe a distinct relation between the two variable, we can observe that the Health Expenditure (% of GDP) variable also show less fluctuation across the years for multiple countries. Lets see if Health Expenditure Per Capita (USD) over the Years shows any reation with annual population growth.

@@

The Health Expenditure Per Capita (USD) varriable in fact shows more variation and also show some connections with annual population growth.
```{r}

### 2025-04-04 products/manuscript/Manuscript.qmd
changes to regression tables

We ran the regression based on our primary model described in the [proposed model for the research](#model) section of this article. The results obtained from the regression analysis are presented by @tbl-preliminary-regression-results.

```{r}
#| label: tbl-preliminary-regression-results
#| tbl-cap: "Preliminary Regression Results"
#| echo: FALSE
#| out.width: "70%"
knitr::include_graphics(here("results", "tables", "preliminary_regression_results.png"))
```

>>>>

We ran the regression based on our primary model described in the [proposed model for the research](#model) section of this article. The results obtained from the regression analysis are presented by @tbl-primary-regression-results.

```{r}
#| label: tbl-primary-regression-results
#| tbl-cap: "Primary Regression Results"
#| echo: FALSE
#| out.width: "70%"
knitr::include_graphics(here("results", "tables", "primary_regression_results.png"))
```
